---
title: "Testing for species niche shifts"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, tidy = FALSE)
```

## Change log

* 13 May: update distributions in regression fits to work with circular data.
* 6 May: change mean time to median time analysis, and use circular.
* 4 May: add winter-only analysis.
* 22 April: first version.

## Summary

Here, I calculated the pairwise distance between the foraging time distributions for all chambers, *within* species at each site. We can test individual pairs of chambers to see if the temporal niche distributions of the chambers are significantly different from each other, but that does not really address the hypothesis. What I did here is run a Mantel test, which is a statistical test to see whether two distance matrices are correlated. The question it answers here is: for each individual ant species, if two chambers are more similar in temperature, is the temporal niche distribution of the ants in those chambers more similar? I used two different measures of distance, one is 1 minus the overlap, the other is the "earth mover's distance" which is another way of looking at the distance between two distributions. In both cases I used circular statistics, and did a separate test for each species. 

I also used the circular median to see whether the average time of activity (not caring at all about the shape of the distribution) changes with temperature. I calculated the median for all the species and chambers and did a single mixed model with all the species.

For the distance tests, basically the answer is no trend. The only one that might be changing with temperature is *Aphaenogaster rudis* at Duke Forest. It has $p < 0.1$ for both distance metrics. It appears to be changing from morning to evening activity as temperature increases. However it's not too surprising to get one nonzero trend from chance alone if you look at enough species. 

For the median tests, it looks like *Aphaenogaster* is getting later (credible interval is greater than zero) and *Prenolepis* is getting earlier (maybe, but the credible interval still overlaps zero) at Duke Forest. So overall I think the hypothesis of little change is supported though there are some potential things happening.

I repeated all the analyses with summer and winter split up, but not a lot was revealed from doing that. At best, 

The raw R code used to generate this document is not in the PDF but it is in the GitHub repository at `github.com/qdread/anttime/nicheshifttests.Rmd`. You can reproduce this analysis by running that code. Note that the `Ostats` package needs to be installed using `devtools::install_github('NEON-biodiversity/Ostats')`.

The histograms for all the different species are not in this document, they are in another document (`time_plots_allspp.pdf`).

```{r load and wrangle data, message = FALSE}
library(tidyverse)
library(ade4)
library(Ostats)
library(emdist)
library(circular)
library(brms)
library(brmstools)

theme_set(theme_minimal())

trt <- read_csv('data/chamber_treatments.csv')
dat <- read_csv('data/data_allmonths_allsites.csv')

# Create a new column in treatment with the temp and chamber ID
# Order it by the temperature so the plots are in ascending temp order
trt <- trt %>%
  mutate(chamber_temp = paste0('chamber ', chamber, ': +', temperature, 'C')) %>%
  mutate(chamber_temp = factor(chamber_temp, 
                               levels = unique(chamber_temp)[order(site, temperature)]))

# Convert month to ordered factor and join with treatment
dat <- dat %>%
  filter(!spp %in% c('none','unk'), !is.na(spp)) %>%
  mutate(month = factor(month, levels = month.name)) %>%
  left_join(trt)

# Additional data wrangling:
# Convert to long form with one row per individual ant observed.
dat_long <- dat %>%
  group_by(site, month, date, time,  chamber, temperature, chamber_temp) %>%
  group_modify(~ data.frame(sp = rep(.$spp, each = .$number)))

# Filter the data to show only species with at least 100 individuals
# Also define seasons
dat_common <- dat_long %>%
  filter(!is.na(temperature)) %>%
  mutate(season = if_else(month %in% month.name[4:9], 'summer', 'winter')) %>%
  group_by(sp) %>%
  filter(n() >= 100)

# Set color palette for all species
sp_descend <- dat_common %>% summarize(n=n()) %>% arrange(-n) %>% pull(sp)
fill_palette <- scale_fill_manual(values = 
  setNames(c(RColorBrewer::brewer.pal(9, 'Set1'), 'turquoise'), sp_descend))
color_palette <- scale_color_manual(values = 
  setNames(c(RColorBrewer::brewer.pal(9, 'Set1'), 'turquoise'), sp_descend))

```


## Overlap within species

Here I calculate all the pairwise distances between the time distributions for all pairs of chambers for each species. Because there are 12 chambers there are 66 possible pairs at each site. I use two different ways of calculating the distance. The first is 1 - the pairwise overlap of the distributions. The second is the earth mover's distance (EMD). EMD represents the amount of work it would take to convert one distribution to another, if you imagine them as two piles of dirt with different heights and in slightly different places.

I got the code for doing the distances over circular distributions from [Kjetil Halvorsen on StackOverflow.com](https://stats.stackexchange.com/questions/461345/earth-movers-distance-implementation-for-circular-distributions). The densities are just the histograms for each species-chamber combination with 24 bins, one for each hour.

I calculated all pairwise distances for all combinations of species and site. Only use species with >100 individuals and chambers with >5 individuals for that species and where there were at least two different times when ants were seen.

```{r define functions}
# Distance between two hour distributions
hourdist <- function(A, B) sum(pmin(  (A-B)%%24, (B-A)%%24 ) )  

# Manually calculate density from the vector of hours
calc_weight <- function(x) { # a vector of hours
    tab <- table(factor(x,  levels=as.character(0:23)),
                 useNA="ifany")

    dimnames(tab) <- NULL
    weights <- tab / sum(tab)

    mat <- cbind( weights=weights, points=0:23 )
    mat
  }

# Calculate EMD or overlap distance for all pairs (function)
all_pairs_asmatrix <- function(traits, sp, metric = c('overlap', 'emd')) {
  require(emdist)
  dat <- data.frame(traits=traits, sp=sp, stringsAsFactors = FALSE)
  dat <- dat[complete.cases(dat), ]
  abunds <- table(dat$sp)
  abunds <- abunds[abunds>1]
  spp <- names(abunds)
  dat <- dat[dat$sp %in% spp, ]
  traitlist <- split(dat$traits, dat$sp)
  nspp <- length(traitlist)
  
  distances <- matrix(NA, nrow = nspp, ncol = nspp)
  
  for (sp_a in 1:(nspp-1)) {
    for (sp_b in (sp_a+1):nspp) {
      a <- traitlist[[sp_a]]
      b <- traitlist[[sp_b]]
      density_a <- calc_weight(a)
      density_b <- calc_weight(b)
      if (metric[1] == 'emd') {
        distances[sp_a, sp_b] <- emd(density_a, density_b, dist = hourdist)       
      } 
      if (metric[1] == 'overlap') {
        distances[sp_a, sp_b] <- 
          1 - circular_overlap_24hour(a, b)[1]
      }
    }
  }
  
  dimnames(distances) <- list(spp, spp)
  distances
}

```

```{r calculate distances}
# Create temperature distance matrix
duke_temp_dist <- dist(trt$temperature[trt$site == 'Duke'])
harvard_temp_dist <- dist(trt$temperature[trt$site == 'Harvard'])

overlap_mats <- dat_common %>%
  group_by(site, sp, chamber) %>%
  filter(n() >= 5 & length(unique(time)) > 1) %>%
  ungroup %>%
  group_by(site, sp) %>%
  filter(length(unique(chamber)) > 1) %>%
  nest %>%
  mutate(temp_dist = if_else(site == 'Duke', 
                             list(duke_temp_dist), 
                             list(harvard_temp_dist))) %>%
  mutate(mat = map(data, 
                   ~ all_pairs_asmatrix(traits = .$time, 
                                        sp = .$chamber, 
                                        metric = 'overlap')),
         mat_emd = map(data, 
                       ~ all_pairs_asmatrix(traits = .$time, 
                                            sp = .$chamber, 
                                            metric = 'emd')),
         temp_dist = map2(mat, temp_dist, 
                          ~ as.dist(as.matrix(.y)[as.numeric(row.names(.x)), 
                                                  as.numeric(row.names(.x))])),
         dist_overlap = map(mat, ~ as.dist(t(.))),
         dist_emd = map(mat_emd, ~ as.dist(t(.))))
```

Here are a few "heat map" plots that attempt to show the (lack of) relationship. If a species' distance matrix is correlated with the temperature distance matrix, it will look more similar. For the temperature heatmap, pairs of chambers with very far apart temperatures are shown in brighter color (yellow) and are at the upper right. If a species would respond in a predictable way to temperature, we would also see more yellow in the corner and more purple on the diagonal. I don't really see it!

I only plot a couple of the species for the sake of space.

```{r define heatmap functions}
# Key to sort matrices by increasing temp.
duke_key <- with(trt[trt$site == 'Duke', ], chamber[order(temperature)])
harvard_key <- with(trt[trt$site == 'Harvard', ], chamber[order(temperature)])

heat_map <- function(mat, key, title) {
  if (class(mat) == 'matrix') mat <- pmin(mat,t(mat), na.rm = TRUE) else mat <- as.matrix(mat)
  mat_sorted <- mat[key, key]
  mat_sorted[lower.tri(mat_sorted, diag = TRUE)] <- NA
  
  rotate <- function(x) t(apply(x, 2, rev))
  
  image(rotate(mat_sorted), 
        xaxt = 'none', yaxt = 'none', bty = 'n', 
        col = hcl.colors(12, palette = 'viridis', rev = FALSE), main = title)
  axis(3, at = seq(0,1,length.out=12)[-1], labels = key[-1], tick = FALSE)
  axis(4, at = seq(1,0,length.out=12)[-12], labels = key[-12], tick = FALSE, las = 1)
}
```


### *Aphaenogaster rudis* (Duke)

```{r apru heatmap}
par(mfcol = c(1,3), pty = 's', mar = c(5.1, 4.1, 6, 2.1))

heat_map(duke_temp_dist/max(duke_temp_dist), 
         duke_key, 'temperature treatments')
heat_map(overlap_mats$mat_emd[[6]]/max(overlap_mats$mat_emd[[6]], na.rm = TRUE), 
         duke_key, 'earth mover\'s distance')
heat_map(overlap_mats$mat[[6]]/max(overlap_mats$mat[[6]], na.rm = TRUE), 
         duke_key, '1 - overlap')

```

### *Prenolepis imparis* (Duke)

```{r prim heatmap}
par(mfcol = c(1,3), pty = 's', mar = c(5.1, 4.1, 6, 2.1))

heat_map(duke_temp_dist/max(duke_temp_dist), 
         duke_key, 'temperature treatments')
heat_map(overlap_mats$mat_emd[[1]]/max(overlap_mats$mat_emd[[1]], na.rm = TRUE), 
         duke_key, 'earth mover\'s distance')
heat_map(overlap_mats$mat[[1]]/max(overlap_mats$mat[[1]], na.rm = TRUE), 
         duke_key, '1 - overlap')

```

### *Crematogaster lineolata* (Duke)

```{r crli heatmap}
par(mfcol = c(1,3), pty = 's', mar = c(5.1, 4.1, 6, 2.1))

heat_map(duke_temp_dist/max(duke_temp_dist), 
         duke_key, 'temperature treatments')
heat_map(overlap_mats$mat_emd[[3]]/max(overlap_mats$mat_emd[[3]], na.rm = TRUE), 
         duke_key, 'earth mover\'s distance')
heat_map(overlap_mats$mat[[3]]/max(overlap_mats$mat[[3]], na.rm = TRUE), 
         duke_key, '1 - overlap')

```

## Mantel tests

The Mantel test is a statistical test to see if two distance matrices are correlated. We do one for each species, where one matrix is the temperature treatment difference between the pairs of chambers and the other is the distance between the foraging time distributions between the pairs of chambers. The alternative hypothesis is that as the difference between the temperatures of two chambers gets greater, the distance between the foraging time distributions will also get greater.

```{r mantel tests}
set.seed(919)

# Mantel tests for the two distance metrics for all species/site combos.
overlap_mats <- overlap_mats %>%
  mutate(mantel_test_overlap = 
           map2(temp_dist, dist_overlap, ~ mantel.rtest(.x, .y, nrepet = 9999)),
         mantel_test_emd = 
           map2(temp_dist, dist_emd, ~ mantel.rtest(.x, .y, nrepet = 9999))
  )

# Extract p values (OMG p values)
overlap_mats <- overlap_mats %>%
  mutate_at(vars(starts_with('mantel_test')), 
            list(stat = ~ map_dbl(., 'obs'), 
                 p_val = ~ map_dbl(., 'pvalue')))
```

If we just look at the p-values, we can see that there is only a marginal trend for *Aphaenogaster rudis* at Duke Forest ($p < 0.1$ for both metrics). The two metrics seem to agree pretty well which is good. *Crematogaster* also has p-values of around 0.1 for both metrics. (*Temnothorax curvispinosus* shows marginal significance for one metric as well but it has fairly few individuals so might be more like noise).

```{r show p-values, echo = FALSE}
overlap_mats %>%
  select(site, sp, contains('p_val'))
```


## Change in median foraging time

It is a little hard to interpret the results in terms of distributions changing. That does not really say anything about what direction they are changing in. To make a more interpretable result, I also calculated the "circular median" of the activity times for each species in each chamber. Then I did a mixed model for each site to see if temperature predicts the central tendency activity time, with species as a random effect (random intercept and random slope). 
We can ask: is the median foraging time getting earlier (or later) with increasing temperature? The random efects model allows us to look at trends for each species while still preserving the non-independence of the species in the same chambers. I am using a Bayesian model with a special prior distribution that works with circular data.

```{r model median foraging time}
mean_times <- dat_common %>%
  group_by(site, sp, chamber, temperature, chamber_temp) %>%
  summarize(median_time = median(circular(time, units = 'hours', modulo = '2pi')))

# Convert median times to radians, from -pi to +pi
mean_times <- mean_times %>%
  mutate(median_time_radians = conversion.circular(median_time, units = 'radians', zero = pi) %>% as.numeric)

options(mc.cores = 2)

# Set prior tighter around intercept (exclude for now)
#tight_prior <- set_prior("student_t(3, 0, 2)", class = "Intercept")

duke_median_bayesfit <- brm(median_time_radians ~ temperature + (temperature | sp), 
                            family = von_mises(), 
                            data = mean_times %>% filter(site == 'Duke'),
                            control = list(adapt_delta = 0.9),
                            chains = 2, iter = 12500, warmup = 10000, seed = 44444)
harvard_median_bayesfit <- brm(median_time_radians ~ temperature + (temperature | sp), 
                               family = von_mises(), 
                               data = mean_times %>% filter(site == 'Harvard'),
                               chains = 2, iter = 12500, warmup = 10000, seed = 55555)
```

These plots show the trends for each species from the mixed model.

```{r panel plots}
# Back-transformation of the radians to hours
hour_y <- scale_y_continuous(name = 'time', breaks = c(-pi, -pi/2, 0, pi/2, pi), labels = c(0, 6, 12, 18, 24))
cutoff_y <- coord_cartesian(ylim = c(-pi,pi))
panels(duke_median_bayesfit, grouping = 'sp', xvar = 'temperature') + 
  ggtitle('Duke Forest') + hour_y + cutoff_y
panels(harvard_median_bayesfit, grouping = 'sp', xvar = 'temperature') + 
  ggtitle('Harvard Forest') + hour_y + cutoff_y

```


These are the coefficients for each species at Duke Forest and at Harvard Forest. The R-squared for the Duke model is `r round(bayes_R2(duke_median_bayesfit)[1,1], 3)` and for the Harvard model it is `r round(bayes_R2(harvard_median_bayesfit)[1,1], 3)`.

```{r show model result, message=TRUE}
message('Duke Forest')
coef(duke_median_bayesfit)$sp[,,'temperature']
message('Harvard Forest')
coef(harvard_median_bayesfit)$sp[,,'temperature']
```

The thing we care about is not really the overall effect of temperature &mdash; the slopes for each individual species are what matter. The 95% credible interval of the slopes for almost all of the individual species overlaps zero (those species have no trend in median activity time with temperature). The only one that does not overlap zero is *A. rudis* at Duke, which is moving from early to late. *Crematogaster* is pretty close to having a positive trend as well. If you look at the plot for *Crematogaster* you can see that the species trend appears to be a bad fit but that's really mainly because a few of the points cross over midnight. It looks like *Crematogaster* may be going from day to middle of the night with warming.

I'm not really sure whether it means too much to look at the median time, since species could have multiple peaks of activity time per day. We will have to think about the interpretation. However it is good that the two species with the most significant trends in our distribution tests are also the two that pop out in our median test.

## Repeat with summer only for Duke

To check whether it's okay to lump together summer and winter for Duke Forest, I ran the exact same tests with summer only. The p-values are shown below.

```{r analysis for summer only, echo = FALSE}
overlap_mats_summer <- dat_common %>%
  filter(site == 'Duke', season == 'summer') %>%
  group_by(site, sp, chamber) %>%
  filter(n() >= 5 & length(unique(time)) > 1) %>%
  ungroup %>%
  group_by(site, sp) %>%
  filter(length(unique(chamber)) > 1) %>%
  nest %>%
  mutate(temp_dist = if_else(site == 'Duke', 
                             list(duke_temp_dist), 
                             list(harvard_temp_dist))) %>%
  mutate(mat = map(data, 
                   ~ all_pairs_asmatrix(traits = .$time, 
                                        sp = .$chamber, 
                                        metric = 'overlap')),
         mat_emd = map(data, 
                       ~ all_pairs_asmatrix(traits = .$time, 
                                            sp = .$chamber, 
                                            metric = 'emd')),
         temp_dist = map2(mat, temp_dist, 
                          ~ as.dist(as.matrix(.y)[as.numeric(row.names(.x)), 
                                                  as.numeric(row.names(.x))])),
         dist_overlap = map(mat, ~ as.dist(t(.))),
         dist_emd = map(mat_emd, ~ as.dist(t(.))))

set.seed(410)

# Mantel tests for the two distance metrics for all species/site combos.
overlap_mats_summer <- overlap_mats_summer %>%
  mutate(mantel_test_overlap = 
           map2(temp_dist, dist_overlap, ~ mantel.rtest(.x, .y, nrepet = 9999)),
         mantel_test_emd = 
           map2(temp_dist, dist_emd, ~ mantel.rtest(.x, .y, nrepet = 9999))
  )

# Extract p values (OMG p values)
overlap_mats_summer <- overlap_mats_summer %>%
  mutate_at(vars(starts_with('mantel_test')), 
            list(stat = ~ map_dbl(., 'obs'), 
                 p_val = ~ map_dbl(., 'pvalue')))

overlap_mats_summer %>%
  select(site, sp, contains('p_val'))
```

We now see a more suggestive trend with *Crematogaster* but less of a trend for the other species. If you look at the histograms for *Crematogaster* it does sort of look like it is less active during the middle of the day in the hotter chambers.

Let's look at the mixed model for median times for summer only too.

```{r model mean time summer only, echo = FALSE}
mean_times_summer <- dat_common %>%
  group_by(site, sp, season, chamber, temperature, chamber_temp) %>%
  summarize(median_time = median(circular(time, units = 'hours', modulo = '2pi')))
  
  # Convert median times to radians, from -pi to +pi
mean_times_summer <- mean_times_summer %>%
  mutate(median_time_radians = conversion.circular(median_time, units = 'radians', zero = pi) %>% as.numeric)

dukesummer_median_bayesfit <- brm(median_time_radians ~ temperature + (temperature | sp), 
                                  family = von_mises(), 
                                  data = mean_times_summer %>% filter(site == 'Duke', season == 'summer'),
                                  control = list(adapt_delta = 0.9),
                                  chains = 2, iter = 12500, warmup = 10000, seed = 66666)

coef(dukesummer_median_bayesfit)$sp[,,'temperature']
```

The trend for *Aphaenogaster rudis* is still the only one that does not overlap zero. *Crematogaster* is no longer close to having a trend.

## Repeat with winter only for Duke

These are the p-values for winter only. *Crematogaster* has the most suggestive trend.

```{r analysis for winter only, echo = FALSE}
overlap_mats_winter <- dat_common %>%
  filter(site == 'Duke', season == 'winter') %>%
  group_by(site, sp, chamber) %>%
  filter(n() >= 5 & length(unique(time)) > 1) %>%
  ungroup %>%
  group_by(site, sp) %>%
  filter(length(unique(chamber)) > 1) %>%
  nest %>%
  mutate(temp_dist = if_else(site == 'Duke', 
                             list(duke_temp_dist), 
                             list(harvard_temp_dist))) %>%
  mutate(mat = map(data, 
                   ~ all_pairs_asmatrix(traits = .$time, 
                                        sp = .$chamber, 
                                        metric = 'overlap')),
         mat_emd = map(data, 
                       ~ all_pairs_asmatrix(traits = .$time, 
                                            sp = .$chamber, 
                                            metric = 'emd')),
         temp_dist = map2(mat, temp_dist, 
                          ~ as.dist(as.matrix(.y)[as.numeric(row.names(.x)), 
                                                  as.numeric(row.names(.x))])),
         dist_overlap = map(mat, ~ as.dist(t(.))),
         dist_emd = map(mat_emd, ~ as.dist(t(.))))

set.seed(411)

# Mantel tests for the two distance metrics for all species/site combos.
overlap_mats_winter <- overlap_mats_winter %>%
  mutate(mantel_test_overlap = 
           map2(temp_dist, dist_overlap, ~ mantel.rtest(.x, .y, nrepet = 9999)),
         mantel_test_emd = 
           map2(temp_dist, dist_emd, ~ mantel.rtest(.x, .y, nrepet = 9999))
  )

# Extract p values (OMG p values)
overlap_mats_winter <- overlap_mats_winter %>%
  mutate_at(vars(starts_with('mantel_test')), 
            list(stat = ~ map_dbl(., 'obs'), 
                 p_val = ~ map_dbl(., 'pvalue')))

overlap_mats_winter %>%
  select(site, sp, contains('p_val'))
```

These are the coefficient estimates from the regression on the median times in winter. For some reason they are giving a weird result. I do not think there are enough data points to fit the model in winter.

```{r model mean time winter only, echo = FALSE}
# Use only the species that have valid overlaps
spp_to_use <- c('prim','crli','apru','apla','caca')


dukewinter_median_bayesfit <- brm(median_time_radians ~ temperature + (temperature | sp), 
                                  family = von_mises(), 
                                  data = mean_times_summer %>% filter(site == 'Duke', season == 'winter', sp %in% spp_to_use),
                                  control = list(adapt_delta = 0.9),
                                  chains = 2, iter = 12500, warmup = 10000, seed = 11111)

coef(dukewinter_median_bayesfit)$sp[,,'temperature']
```