---
title: "Foraging by time plots by site, species, and season"
author: "Quentin D. Read"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

# Contents

These are the ridge plots, created with `ggridges` package, for all the species and sites (7 spp. at Duke and 3 spp. at Harvard with adequate sample sizes). Separate ones are made for summer and winter, then there is one for each species with both seasons combined. The plots are also written to files.


```{r load and wrangle data}
source('load_wrangle_data.R')
library(ggridges) # For ridgeline plot


# Define seasons
dat_common <- dat_common %>%
  mutate(season = if_else(month %in% month.name[4:9], 'summer', 'winter')) 

# Add dummy data so all panels will plot
dummy_data <- dat_common %>%
  group_by(site,season,chamber_temp,sp) %>%
  slice(1) %>%
  mutate(time = -1) %>% 
  ungroup

sp_season <- expand_grid(sp = sp_descend, season = c('summer','winter'))
dummy_data <- dat_common %>% 
  ungroup %>% 
  select(site, chamber_temp) %>% 
  unique %>%
  group_by(site, chamber_temp) %>% group_modify(~ data.frame(sp_season))


dat_common_toplot <- bind_rows(dat_common, dummy_data)

temp_fill_palette <- scale_fill_distiller(palette = 'RdYlBu')
```

```{r calculate circular means and CIs}
# Make sure not to use the one with dummy data included.

summary_stats <- dat_common %>%
  group_by(site, sp, chamber, temperature, season, chamber_temp) %>%
  summarize(mean_time = mean(circular(time, units = 'hours', modulo = '2pi')),
            q05_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.05),
            q25_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.25),
            median_time = median(circular(time, units = 'hours', modulo = '2pi')),
            q75_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.75),
            q95_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.95)
            )

# Include code to split up the segment into multiple chunks if it goes across midnight.
# For the 25 to 75 quantile segment, this will be true if q25 > q75.

summary_stats <- summary_stats %>%
  mutate(seg1_start = if_else(q25_time > q75_time, 0, as.numeric(q25_time)),
         seg1_end = if_else(q25_time > q75_time, as.numeric(q75_time), as.numeric(q75_time)),
         seg2_start = if_else(q25_time > q75_time, as.numeric(q25_time), as.numeric(NA)),
         seg2_end = if_else(q25_time > q75_time, 23, as.numeric(NA)))

### Summary stats not split by season.
summary_stats_combineseasons <- dat_common %>%
  group_by(site, sp, chamber, temperature,  chamber_temp) %>%
  summarize(mean_time = mean(circular(time, units = 'hours', modulo = '2pi')),
            q05_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.05),
            q25_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.25),
            median_time = median(circular(time, units = 'hours', modulo = '2pi')),
            q75_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.75),
            q95_time = quantile(circular(time, units = 'hours', modulo = '2pi'), probs = 0.95)
  ) %>%
  mutate(seg1_start = if_else(q25_time > q75_time, 0, as.numeric(q25_time)),
         seg1_end = if_else(q25_time > q75_time, as.numeric(q75_time), as.numeric(q75_time)),
         seg2_start = if_else(q25_time > q75_time, as.numeric(q25_time), as.numeric(NA)),
         seg2_end = if_else(q25_time > q75_time, 23, as.numeric(NA)))

```


```{r define functions}
two_season_plot <- function(site_to_plot, sp_to_plot) {
  plotdat <- dat_common_toplot %>% filter(site == site_to_plot, sp == sp_to_plot)
  plotstats <- summary_stats %>% filter(site == site_to_plot, sp == sp_to_plot)
  ggplot(plotdat, 
         aes(x = time, y= chamber_temp, fill = temperature)) +
    geom_density_ridges(aes(y = chamber_temp), alpha = 0.7, color = 'gray50', from = 0, to = 23, bandwidth = 1) +
    geom_segment(aes(x = seg1_start, xend = seg1_end, yend = chamber_temp), data = plotstats, size = 1.5) +
    geom_segment(aes(x = seg2_start, xend = seg2_end, yend = chamber_temp), data = plotstats, size = 1.5) +
    geom_point(aes(x = median_time), data = plotstats, size = 2, fill = 'gray50', shape = 21, stroke = 2) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_discrete(expand = c(0,0)) +
    temp_fill_palette +
    facet_wrap(~ season) +
    coord_cartesian(clip = 'off') +
    theme_ridges(grid = FALSE) +
    theme(legend.position = 'none', 
          axis.title.y = element_blank(),
          strip.background = element_blank(),
          panel.background = element_rect(colour = 'black', size = 0.75, linetype = 1),
          panel.spacing = unit(2, 'lines'))
}

one_season_plot <- function(site_to_plot, sp_to_plot, season_to_plot) {
  plotdat <- dat_common_toplot %>% filter(site == site_to_plot, sp == sp_to_plot, season %in% season_to_plot)
  plotstats <- summary_stats %>% filter(site == site_to_plot, sp == sp_to_plot, season %in% season_to_plot)
  ggplot(plotdat, 
         aes(x = time, y= chamber_temp, fill = temperature)) +
    geom_density_ridges(aes(y = chamber_temp), alpha = 0.7, color = 'gray50', from = 0, to = 23, bandwidth = 1) +
    geom_segment(aes(x = seg1_start, xend = seg1_end, yend = chamber_temp), data = plotstats, size = 1.5) +
    geom_segment(aes(x = seg2_start, xend = seg2_end, yend = chamber_temp), data = plotstats, size = 1.5) +
    geom_point(aes(x = median_time), data = plotstats, size = 2, fill = 'gray50', shape = 21, stroke = 2) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_discrete(expand = c(0,0)) +
    temp_fill_palette +
    coord_cartesian(clip = 'off') +
    theme_ridges(grid = FALSE) +
    theme(legend.position = 'none', 
          axis.title.y = element_blank())
}

combined_season_plot <- function(site_to_plot, sp_to_plot, season_to_plot) {
  plotdat <- dat_common_toplot %>% filter(site == site_to_plot, sp == sp_to_plot, season %in% season_to_plot)
  plotstats <- summary_stats_combineseasons %>% filter(site == site_to_plot, sp == sp_to_plot)
  ggplot(plotdat, 
         aes(x = time, y= chamber_temp, fill = temperature)) +
    geom_density_ridges(aes(y = chamber_temp), alpha = 0.7, color = 'gray50', from = 0, to = 23, bandwidth = 1) +
    geom_segment(aes(x = seg1_start, xend = seg1_end, yend = chamber_temp), data = plotstats, size = 1.5) +
    geom_segment(aes(x = seg2_start, xend = seg2_end, yend = chamber_temp), data = plotstats, size = 1.5) +
    geom_point(aes(x = median_time), data = plotstats, size = 2, fill = 'gray50', shape = 21, stroke = 2) +
    scale_x_continuous(expand = c(0,0)) +
    scale_y_discrete(expand = c(0,0)) +
    temp_fill_palette +
    coord_cartesian(clip = 'off') +
    theme_ridges(grid = FALSE) +
    theme(legend.position = 'none', 
          axis.title.y = element_blank())
}

```

```{r create plot objects}
harvard_spp <- c('cape', 'apru', 'fosu')

duke_summer_plots <- map(sp_descend, ~ one_season_plot("Duke", ., "summer"))
duke_winter_plots <- map(sp_descend, ~ one_season_plot("Duke", ., "winter"))
duke_bothseasons_plots <- map(sp_descend, ~ combined_season_plot("Duke", ., c("summer", "winter")))
harvard_summer_plots <- map(harvard_spp, ~ one_season_plot("Harvard", ., "summer"))

duke_names <- gsub(' ', '_', spp$species[match(sp_descend, spp$sp)])
harvard_names <- gsub(' ', '_', spp$species[match(harvard_spp, spp$sp)])
```

# Duke Forest

## Prenolepis imparis

```{r duke 1}
print(duke_summer_plots[[1]] + ggtitle('summer'))
print(duke_winter_plots[[1]] + ggtitle('winter'))
print(duke_bothseasons_plots[[1]] + ggtitle('both seasons combined'))
```

## Crematogaster lineolata

```{r duke 2}
print(duke_summer_plots[[2]] + ggtitle('summer'))
print(duke_winter_plots[[2]] + ggtitle('winter'))
print(duke_bothseasons_plots[[2]] + ggtitle('both seasons combined'))
```

## Camponotus pennsylvanicus

```{r duke 3}
print(duke_summer_plots[[3]] + ggtitle('summer'))
print(duke_winter_plots[[3]] + ggtitle('winter'))
print(duke_bothseasons_plots[[3]] + ggtitle('both seasons combined'))
```

## Camponotus castaneus

```{r duke 4}
print(duke_summer_plots[[4]] + ggtitle('summer'))
print(duke_winter_plots[[4]] + ggtitle('winter'))
print(duke_bothseasons_plots[[4]] + ggtitle('both seasons combined'))
```

## Aphaenogaster rudis

```{r duke 5}
print(duke_summer_plots[[5]] + ggtitle('summer'))
print(duke_winter_plots[[5]] + ggtitle('winter'))
print(duke_bothseasons_plots[[5]] + ggtitle('both seasons combined'))
```

## Formica subsericea

```{r duke 6}
print(duke_summer_plots[[6]] + ggtitle('summer'))
print(duke_winter_plots[[6]] + ggtitle('winter'))
print(duke_bothseasons_plots[[6]] + ggtitle('both seasons combined'))
```

## Aphaenogaster lamellidens

```{r duke 7}
print(duke_summer_plots[[7]] + ggtitle('summer'))
print(duke_winter_plots[[7]] + ggtitle('winter'))
print(duke_bothseasons_plots[[7]] + ggtitle('both seasons combined'))
```

\newpage

# Harvard Forest

Only summer observations were done at Harvard Forest.

## Camponotus pennsylvanicus

```{r harvard 1}
print(harvard_summer_plots[[1]])
```

## Aphaenogaster rudis

```{r harvard 2}
print(harvard_summer_plots[[2]])
```

## Formica subsericea

```{r harvard 3}
print(harvard_summer_plots[[3]])
```
